---
title: "Data Cleaning"
author: "Andrew Jaffe"
date: "January 6, 2016"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: ../../styles.css
---

## Data

* We will be using multiple data sets in this lecture:
    * Salary, Monument, Circulator, and Restaurant from OpenBaltimore: 
    https://data.baltimorecity.gov/browse?limitTo=datasets
    * Gap Minder - very interesting way of viewing longitudinal data
        * Data is here - http://www.gapminder.org/data/
    * [http://spreadsheets.google.com/pub?key=rMsQHawTObBb6_U2ESjKXYw&output=xls](http://spreadsheets.google.com/pub?key=rMsQHawTObBb6_U2ESjKXYw&output=xls)

## Data Cleaning

In general, data cleaning is a process of investigating your data for inaccuracies, or recoding it in a way that makes it more manageable.

MOST IMPORTANT RULE - LOOK AT YOUR DATA!

Again - `table`, `summarize`, `is.na`, `any`, `all` are useful.


## Missing data
One of the most important aspects of data cleaning is missing values.  

Types of "missing" data:

* `NA` - general missing data
* `NaN` - stands for "**N**ot **a** **N**umber", happens when you do 0/0.
* `Inf` and `-Inf` - Infinity, happens when you take a positive number (or negative number) by 0.

## Finding Missing data

Each missing data type has a function that returns `TRUE` if the data is missing:

* `NA` - `is.na`
* `NaN` - `is.nan`
* `Inf` and `-Inf` - `is.infinite`
* `is.finite` returns `FALSE` for all missing data and `TRUE` for non-missing
* `complete.cases` on a `data.frame`/`matrix` returns `TRUE` if all values in that row of the object are not missing.

## Missing Data with Logicals

One important aspect (esp with subsetting) is that logical operations return `NA` for `NA` values.  Think about it, the data could be `> 2` or not we don't know, 
so `R` says there is no `TRUE` or `FALSE`, so that is missing:
```{r}
x = c(0, 1, 2, 3, NA, 3, 3, 2,2, 3)
x > 2
```

## Missing Data with Logicals

What to do?  What if we want if `x > 2` and `x` isn't `NA`?  
Don't do `x != NA`, do `x > 2` and `x` is NOT `NA`:

```{r}
x != NA
x > 2 & !is.na(x)
```

## Missing Data with Logicals

What about seeing if a value is equal to multiple values?  You can do `(x == 1 | x == 2) & !is.na(x)`, but that is not efficient.  Introduce the `%in%` operator:

```{r}
(x == 1 | x == 2) # has NA
(x == 1 | x == 2) & !is.na(x) # No NA
x %in% c(1, 2) # NEVER has NA and returns logical
```

## Creating One-way Tables {.smaller}

Here we will use `table` to make tabulations of the data.   Look at `?table` to see options for missing data.

```{r table}
table(x)
table(x, 
        useNA = "ifany")
```


## Creating One-way Tables {.smaller}

You can set `useNA = "always"` to have it always have a column for `NA`
```{r onetab}
table(c(0, 1, 2, 3, 2, 3, 3, 2,2, 3), 
        useNA = "always")
```

## Data Cleaning {.smaller}

A two-way table.  If you pass in 2 vectors, `table` creates a 2-dimensional table.

```{r}

tab <- table(c(0, 1, 2, 3, 2, 3, 3, 2,2, 3), 
             c(0, 1, 2, 3, 2, 3, 3, 4, 4, 3), 
              useNA = "always")
```

## Finding Row or Column Totals

`margin.table` finds the marginal sums of the table.  `margin` is 1 for rows, 2 for columns in general in `R`.  Here is the column sums of the table:

```{r margin}
margin.table(tab, 2)
```

## Data Cleaning {.smaller}

`prop.table` finds the marginal proportions of the table. Think of it dividing the table by it's respective marginal totals.  If `margin` not set, divides by overall total.

```{r table2}
prop.table(tab)
prop.table(tab,1)
```

## Download Salary FY2014 Data

From https://data.baltimorecity.gov/City-Government/Baltimore-City-Employee-Salaries-FY2014/2j28-xzd7
http://www.aejaffe.com/winterR_2016/data/Baltimore_City_Employee_Salaries_FY2014.csv

Read the CSV into R `Sal`

```{r readSal}
Sal = read.csv("http://www.aejaffe.com/winterR_2016/data/Baltimore_City_Employee_Salaries_FY2014.csv",
               as.is = TRUE)
```

## Data Cleaning {.smaller}
* `any()` - checks if there are any `TRUE`s
* `all()` - checks if ALL are true

```{r isna}
Sal[1:4,]
any(is.na(Sal$Name)) # are there any NAs?
```

## Example of Cleaning:

For example, let's say gender was coded as Male, M, m, Female, F, f.  Using Excel to find all of these would be a matter of filtering and changing all by hand or using if statements.

In R, you can simply do something like:

```
data$gender[data$gender %in% 
    c("Male", "M", "m")] <- "Male"
```

## Example of Cleaning:

You can also `recode` a vector:

```{r}
library(car)
x = rep(c("Male", "M", "m", "f", "Female", "female" ), each = 3)
car::recode(x, "c('m', 'M', 'male') = 'Male';
            c('f', 'F', 'female') = 'Female';")
```

## Example of Cleaning:

You can also `revalue` a vector with the `revalue` command

```{r}
library(plyr)
plyr::revalue(x, c("M" = "Male", "m" = "Male",
                   "f" = "Female", "female" = "Female"))
```

## Example of Cleaning:

Sometimes though, it's not so simple.  That's where functions that find patterns come in very useful.

```{r gender, echo=FALSE}
set.seed(4)
gender <- sample(c("Male", "mAle", "MaLe", "M", "MALE", "Ma", "FeMAle", "F", "Woman", "Man", "Fm", "FEMALE"), 1000, replace =TRUE)
```

```{r gentab}
table(gender)
```

## Find/Replace and Regular Expressions

* R can do much more than find exact matches for a whole string
* Like Perl and other languages, it can use regular expressions.
* What are regular expressions?
* Ways to search for specific strings 
* Can be very complicated or simple
* Highly Useful

## 'Find' functions

`grep`: `grep`, `grepl`, `regexpr` and `gregexpr` search for matches to argument pattern within each element of a character vector: they differ in the format of and amount of detail in the results. 

`grep(pattern, x, fixed=FALSE)`, where:

* pattern = character string containing a regular expression to be matched in the given character vector.

* x = a character vector where matches are sought, or an object which can be coerced by as.character to a character vector.

* If fixed=TRUE, it will do exact matching for the phrase anywhere in the vector (regular find)

## 'Find' functions

```{r RawlMatch}
grep("Rawlings",Sal$Name)
```

These are the indices/elements where the pattern match occurs

`grep()` returns something similar to which() on a logical statement

## 'Find' functions {.smaller}

```{r grepl}
grep("Rawlings",Sal$Name)
grep("Rawlings",Sal$Name,value=TRUE)
Sal[grep("Rawlings",Sal$Name),]
```

## `grep()` Options
```{r greppers}
head(grep("Tajhgh",Sal$Name, value=TRUE))
grep("Jaffe",Sal$Name)
length(grep("Jaffe",Sal$Name))
```

## A bit on Regular Expressions

* http://www.regular-expressions.info/reference.html
* They can use to match a large number of strings in one statement
* `.` matches any single character
* `*` means repeat as many (even if 0) more times the last character
* `?` makes the last thing optional

## Using Regular Expressions
* Look for any name that starts with:
    * Payne at the beginning, 
    * Leonard and then an S
    * Spence then a capital C
  
```{r grepstar}
grep("Payne.*", x=Sal$Name, value=TRUE)
```



```{r grepstar2}
grep("Leonard.?S", x=Sal$Name, value=TRUE)
grep("Spence.*C.*", x=Sal$Name, value=TRUE)
```

## Replace

Let's say we wanted to sort the data set by Annual Salary:

```{r classSal}
class(Sal$AnnualSalary)
```

```{r orderstring}
sort(c("1", "2", "10")) #  not sort correctly (order simply ranks the data)
order(c("1", "2", "10"))
```

## Replace

So we must change the annual pay into a numeric:
```{r destringSal}
head(as.numeric(Sal$AnnualSalary), 4)
```

R didn't like the $ so it thought turned them all to `NA`.

`sub()` and `gsub()` can do the replacing part.

## Replacing and subbing {.smaller}

Now we can replace the `$` with nothing (used fixed=TRUE because $ means something in regular expressions):

```{r orderSal}
Sal$AnnualSalary <- as.numeric(gsub(pattern="$", replacement="", 
                              Sal$AnnualSalary, fixed=TRUE))
Sal <- Sal[order(Sal$AnnualSalary,decreasing=TRUE), ] # use negative to sort descending
Sal[1:5, c("Name", "AnnualSalary", "JobTitle")]
```

## Useful String Functions

Useful String functions

* `toupper()`, `tolower()` - uppercase or lowercase your data:
* `str_trim()` (in the `stringr` package) - will trim whitespace
* `nchar` - get the number of characters in a string
* `substr(x, start, stop)` - substrings from position start to position stop
* `strsplit(x, split)` - splits strings up - returns list!
* `paste()` - paste strings together - look at `?paste`

## Paste

Paste can be very useful for joining vectors together:

```{r Paste}
paste("Visit", 1:5, sep="_")
paste("Visit", 1:5, sep="_", collapse=" ")
paste("To", "is going be the ", "we go to the store!", sep="day ")
# and paste0 can be even simpler see ?paste0 
paste0("Visit",1:5)
```



```{r Paste2}
paste(1:5, letters[1:5], sep="_")
paste(6:10, 11:15, 2000:2005, sep="/")
paste(paste("x",1:5,sep=""),collapse="+")
```

## Strsplit

```{r strsplit}
x <- c("I really", "like writing", "R code")
y <- strsplit(x, split=" ")
y[[2]]
sapply(y, "[", 1) # on the fly
sapply(y, "[", 2) # on the fly
```

## Data Merging/Append

* Merging - joining data sets together - usually on key variables, usually "id"
* `merge()` is the most common way to do this with data sets
* `rbind`/`cbind` - row/column bind, respectively
    * `rbind` is the equivalent of "appending" in Stata or "setting" in SAS
    * `cbind` allows you to add columns in addition to the previous ways
* `reshape2` package also has a lot of information about different ways to reshape data (wide to long, etc) - but has a different (and sometimes more intuitive syntax)
* `t()` is a function that will transpose the data

## Merging {.smaller}
```{r merging}
base <- data.frame(id=1:10, Age= seq(55,60, length=10))
base[1:2,]
visits <- data.frame(id=rep(1:8, 3), visit= rep(1:3, 8),
                    Outcome= seq(10,50, length=24))
visits[1:2,]
```



```{r merging2}
merged.data <- merge(base, visits, by="id")
merged.data[1:5,]
dim(merged.data)
```



```{r mergeall}
all.data <- merge(base, visits, by="id", all=TRUE)
tail(all.data)
dim(all.data)
```

## Aside: Dates {.smaller}

You can convert date-like strings in the `Date` class (http://www.statmethods.net/input/dates.html for more info)

```{r date}
circ = read.csv("http://www.aejaffe.com/winterR_2016/data/Charm_City_Circulator_Ridership.csv",
                as.is=TRUE)
head(sort(circ$date))
circ$date <- as.Date(circ$date, "%m/%d/%Y") # creating a date for sorting
head(circ$date)
head(sort(circ$date))
```

## Data Reshaping

Disclaimer: the `reshape` command in R is not remarkably intuitive.

* Wide - multiple measurements are variables / columns so that the data gets wider with more measurements
* Long - multiple measurements are rows so data gets longer with more measurements
* One example would be many ids with multiple visits

```{r longwide, echo=FALSE}
wide <- data.frame(id=1, visit1="Good", visit2="Good", visit3 = "Bad")
long <- data.frame(id=rep(1, 3), visit=c(1, 2, 3), Outcome=c("Good", "Good", "Bad"))
```

## Example of Long/Wide
```{r showlong, echo=TRUE}
head(wide)
head(long)
```

## Data Reshaping

* Good resource: http://www.ats.ucla.edu/stat/r/faq/reshape.htm

```{r reshape}
head(Indometh) # this is long
```

## Data Reshaping {.smaller}
```{r reshape2}
wide <- reshape(Indometh, v.names = "conc", idvar = "Subject",
                timevar = "time", direction = "wide")
head(wide)
```

## Data Reshaping {.smaller}

```{r reshape3}
dim(Indometh)
wide
```

## Data Reshaping

* If you've reshaped a data set - to get it back, just reshape it again

```{r rewide}
reshape(wide, direction = "long")[1:10,]
```

Note the row name change

## Data Reshaping - A Better Example {.smaller}

```{r xlsx,echo=FALSE, results='hide'}
library(xlsx,verbose=FALSE)
```

```{r TB}
destfile = tempfile(fileext = ".xlsx")
download.file("http://www.aejaffe.com/winterR_2016/data/indicator_estimatedincidencealltbper100000.xlsx", destfile = destfile)
TB <- read.xlsx(file = destfile,
                sheetName = "Data")
head(TB, 1)
TB$NA. <- NULL
head(TB, 1)
```

## Data Reshaping - A Better Example {.smaller}

```{r TB.hd}
colnames(TB) <- c("Country", paste("Year", 
                          1990:2007, sep="."))
head(TB,1)
```

## Data Reshaping - More is better! {.smaller}

```{r TB.long}
TB.long <- reshape(TB, idvar="Country", 
            v.names="Cases", times=1990:2007, 
                   direction="long", timevar="Year", 
                   varying = paste("Year", 1990:2007, sep="."))
head(TB.long, 4)
rownames(TB.long) <- NULL
head(TB.long, 4)
```

