---
title: "Data Input/Output"
author: "Andrew Jaffe"
date: "January 4, 2016"
output:
  ioslides_presentation:
    css: ../../styles.css
  pdf_document:
    toc: false    
---


## Before we get Started: Working Directories

* R looks for files on your computer relative to the "working" directory
* It's always safer to set the working directory at the beginning of your script. Note that setting the working directory created the necessary code that you can copy into your script.
* Example of help file

```{r workingDirectory,eval=FALSE}
## get the working directory
getwd()
# setwd("~/Dropbox/winterR_2016/Lectures")
```

## Setting a Working Directory

* Setting the directory can sometimes be finicky
    * Windows: Default directory structure involves single backslashes ("\"), but R interprets these as "escape" characters. So you must replace the backslash with forward slashed ("/") or two backslashes ("\\")
    * Mac/Linux: Default is forward slashes, so you are okay
* Typical linux/DOS directory structure syntax applies
    * ".." goes up one level
    * "./" is the current directory
    * "~" is your home directory

## Working Directory 

Try some directory navigation: 

```{r directoryNav}
dir("./") # shows directory contents
dir("..")
```

## Working Directory

* Copy the code to set your working directory from the History tab in RStudio (top right)
* Confirm the directory contains "day1.R" using `dir()`


## Data Input

* We used several pre-installed sample datasets during previous modules (`CO2`, `iris`) 
* However, 'reading in' data is the first step of any real project/analysis
* R can read almost any file format, especially via add-on packages
* We are going to focus on simple delimited files first
    * tab delimited (e.g. '.txt')
    * comma separated (e.g. '.csv')
    * Microsoft excel (e.g. '.xlsx')

## Data Input {.smaller}

`read.table()`: Reads a file in table format and creates a data frame from it, with cases corresponding to lines and variables to fields in the file.

```
# the four ones I've put at the top are the important inputs
read.table( file, # filename
           header = FALSE, # are there column names?
           sep = "", # what separates columns?
           as.is = !stringsAsFactors, # do you want character strings as factors or characters?
           quote = "\"'",  dec = ".", row.names, col.names,
           na.strings = "NA", nrows = -1,
           skip = 0, check.names = TRUE, fill = !blank.lines.skip,
           strip.white = FALSE, blank.lines.skip = TRUE, comment.char = "#",
           stringsAsFactors = default.stringsAsFactors())
           
# for example: `read.table("file.txt", header = TRUE, sep="\t", as.is=TRUE)`
```

## Data Input

* The filename is the path to your file, in quotes
* The function will look in your "working directory" if no absolute file path is given
* Note that the filename can also be a path to a file on a website (e.g. 'www.someurl.com/table1.txt')

## Data Aside

* Everything we do in class will be using real publicly available data - there are few 'toy' example datasets and 'simulated' data
* OpenBaltimore and Data.gov will be sources for the first few days

## Data Input

Monuments Dataset: "This data set shows the point location of Baltimore City monuments. However, the completness and currentness of these data are uncertain."

* Download data from http://www.aejaffe.com/winterR_2016/data/Monuments.csv
* Save it (or move it) to the same folder as your day1.R script
* Within RStudio: Session --> Set Working Directory --> To Source File Location
* (data downloaded from https://data.baltimorecity.gov/Community/Monuments/cpxf-kxp3)

## Data Input
R Studio features some nice "drop down" support, where you can run some tasks by selecting them from the toolbar.

For example, you can easily import text datasets using the "Tools --> Import Dataset" command. Selecting this will bring up a new screen that lets you specify the formatting of your text file. 

After importing a datatset, you get the corresponding R commands that you can enter in the console if you want to re-import data.

## Data Input

There is a 'wrapper' function for reading CSV files: 

```{r readCSV}
read.csv
```

Note: the `...` designates extra/optional arguments that can be passed to `read.table()` if needed

## Data Input {.smaller}

* Starting out, you can use RStudio --> Tools --> Import Dataset --> From Text File and select

```{r readCSV2}
mon = read.csv("../../data/Monuments.csv",header=TRUE,as.is=TRUE)
head(mon)
```

## Data Input {.smaller}

```{r subset5}
colnames(mon)
head(mon$zipCode)
head(mon$neighborhood)
```

## Data Input

The `read.table()` function returns a `data.frame`

```{r readCSV3}
class(mon)
str(mon)
```

## Data Input

Changing variable names in `data.frame`s works using the `names()` function, which is analagous to `colnames()` for data frames (they can be used interchangeably)

```{r names1}
names(mon)[1] = "Name"
names(mon)
names(mon)[1] = "name"
names(mon)
```
