---
title: "Manipulating Data in R"
author: "John Muschelli"
date: "January 4, 2016"
output:
  ioslides_presentation:
    css: ../../styles.css
  beamer_presentation: default
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(comment = "")
```

## Overview

In this module, we will show you how to:

1. Perform operations by a grouping variable
2. Reshaping data from long (tall) to wide (fat)
3. Reshaping data from wide (fat) to long (tall)

## Setup

We will show you how to do each operation in base R then show you how to use the `dplyr` or `tidyr` package to do the same operation (if applicable).  

See the "Data Wrangling Cheat Sheet using `dplyr` and `tidyr`":
* https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

## Load the packages/libraries

```{r}
library(dplyr)
library(tidyr)
```

## Data used: Charm City Circulator

http://www.aejaffe.com/winterR_2016/data/Charm_City_Circulator_Ridership.csv

Let's read in the Charm City Circulator data:

```{r}
ex_data = read.csv("http://www.aejaffe.com/winterR_2016/data/Charm_City_Circulator_Ridership.csv", as.is = TRUE)
head(ex_data, 2)
```

## Creating a Date class from a character date

The `lubridate` package is great for dates:

```{r}
library(lubridate) # great for dates!
ex_data = mutate(ex_data, date = mdy(date))
nrow(ex_data[ is.na(ex_data$date), ])
```

## Making column names a little more separated

We will use `str_replace` from `stringr` to put periods in the column names.
```{r}
library(stringr)
cn = colnames(ex_data)
cn = cn %>% 
  str_replace("Board", ".Board") %>% 
  str_replace("Alight", ".Alight") %>% 
  str_replace("Average", ".Average") 
colnames(ex_data) = cn
```


## Removing the daily ridership

We want to look at each ridership, and will removet the `daily` column:
```{r}
ex_data$daily = NULL
```

# Reshaping data from wide (fat) to long (tall)

# Resources

See http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/


## Reshaping data from wide (fat) to long (tall): base R

The `reshape` command exists.  It is a **confusing** function.  Don't use it.

## Reshaping data from wide (fat) to long (tall): tidyr

In `tidyr`, the `gather` function gathers columns into rows.

We want the column names into "`type`" variable in the output dataset and the value in "`number`" variable
```{r}
long = gather(ex_data, "var", "number", 
              starts_with("orange"),
              starts_with("purple"), starts_with("green"),
              starts_with("banner"))
head(long)
table(long$var)
```

## Reshaping data from wide (fat) to long (tall): tidyr

Now each `var` is boardings, averages, or alightings.  We want to separate these so we can have these by line.

```{r}
long = separate_(long, "var", into = c("line", "type"), sep = "[.]")
head(long)
table(long$line)
table(long$type)
```

# Reshaping data from long (tall) to wide (fat)

## Reshaping data from long (tall) to wide (fat): tidyr

In `tidyr`, the `spread` function spreads rows into columns.  Now we have a long data set, but we want to separate the Average, Alightings and Boardings into different columns:

```{r}
# have to remove missing days
wide = filter(long, !is.na(date))
wide = spread(wide, type, number)
head(wide)
```


## Reshaping data from long (tall) to wide (fat): tidyr

We can use `rowSums` to see if any values in the row is `NA` and keep if the row, which is a combination of date and line type has any non-missing data.

```{r}
# wide = wide %>%
#     select(Alightings, Average, Boardings) %>%
#     mutate(good = rowSums(is.na(.)) > 0)
namat = !is.na(select(wide, Alightings, Average, Boardings))
head(namat)
wide$good = rowSums(namat) > 0
head(wide, 3)
```

## Reshaping data from long (tall) to wide (fat): tidyr

Now we can filter only the good rows and delete the `good` column.
```{r}
wide = filter(wide, good) %>% select(-good)
head(wide)
```


# Perform Operations By Groups of Variables

## Perform Operations By Groups: base R

The `tapply` command will take in a vector (`X`), perform a function
(`FUN`) over an index (`INDEX`):

```{r}
args(tapply)
```


## Perform Operations By Groups: base R

Let's get the mean Average ridership by line:
```{r}
tapply(wide$Average, wide$line, mean, na.rm = TRUE)
```

## Perform Operations By Groups: dplyr

Let's get the mean Average ridership by line  We will use `group_by` to group the data by line, then use `summarize` (or `summarise`) to get the mean Average ridership:
```{r}
gb = group_by(wide, line)
summarize(gb, mean_avg = mean(Average))
```

## Perform Operations By Groups: dplyr with piping

Using piping, this is:

```{r}
wide %>% 
  group_by(line) %>%
  summarise(mean_avg = mean(Average))
```


## Perform Operations By Multiple Groups: dplyr 

This can easily be extended using `group_by` with multiple groups.  Let's define the year of riding:

```{r}
wide = wide %>% mutate(year = year(date),
                       month = month(date))
wide %>% 
  group_by(line, year) %>%
  summarise(mean_avg = mean(Average))
```

## Perform Operations By Multiple Groups: dplyr {.smaller}

We can then easily plot each day over time:

```{r}
library(ggplot2)
ggplot(aes(x = date, y = Average, 
               colour = line), data = wide) + geom_line()
```

## Perform Operations By Multiple Groups: dplyr {.smaller}

Let's create the middle of the month (the 15th for example), and name it
mon.

```{r}
mon = wide %>% 
  dplyr::group_by(line, month, year) %>%
  dplyr::summarise(mean_avg = mean(Average))
mon = mutate(mon, 
             mid_month = dmy(paste0("15-", month, "-", year)))
head(mon)
```


## Perform Operations By Multiple Groups: dplyr {.smaller}

We can then easily plot the mean of each month to see a smoother output:

```{r}
ggplot(aes(x = mid_month,
               y = mean_avg, 
               colour = line), data = mon) + geom_line()
```



## Bonus! Points with a smoother! {.smaller}
```{r}
ggplot(aes(x = date, y = Average, colour = line), 
           data = wide) + geom_smooth(se = FALSE) + 
  geom_point(size = .5)
```
